A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-02-19 tex/body.tex
more

Codatatypes are freely generated by their constructors, but in contrast with datatypes,
infinit\-e constructor terms are also legitimate values for codatatypes
(Section~\ref{sec:the-theory-of-co-datatypes}). Intuitively, the
values of a codatatype consist of all well-typed finite and infinite ground
constructor
terms, and only those. For example, the coinductive specification
%
\[\keyw{codatatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat})\]
%
(using an ML-like syntax) introduces a type that
models the natural numbers $\const{Z}$, $\const{S}(\const{Z})$, $\const{S}(\const{S}(\const{Z}))$, $\ldots$\afterLdots{},
using Peano notation, but extended with an
infinite value $\infty = \const{S}(\const{S}(\const{S}(\ldots)))$.
The equation $\const{S}(\infty) \teq \infty$ holds as expected,
because both sides expand to the infinite term
$\const{S}(\const{S}(\const{S}(\ldots)))$, which uniquely identifies the
value~$\infty$.
Compared with the conventional inductive definition
\begin{rep}
\[\keyw{datatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat}) \,\mid\, \const{Infty}\]
\end{rep}
with a dedicated constructor for representing infinity, the codatatype avoids
one case by unifying the finite and infinite nonzero cases.

Datatypes and codatatypes are an integral part of many proof assistants,
including Agda, Coq, Isabelle, Matita, and PVS. In recent years, datatypes
have emerged in a few automatic theorem provers as well. The SMT-LIB
%\cite{barrett-et-al-2010}
format, implemented by most SMT
%(satisfiability modulo theories)
solvers, has been extended with a syntax for datatypes.
In this \thewordpaper, we introduce a unified decision procedure for ground
problems involving datatypes and codatatypes in
\begin{conf}\goodbreak\noindent\end{conf}%  %%% TYPESETTING: terrible HACK
combination (Section~\ref{sec:the-ground-decision-procedure}).
The procedure is described abstractly as a calculus and is
designed to be composable via the Nelson--Oppen method \cite{nelson-oppen-1979}.
It generalizes the procedure by Barrett et al.\ \cite{barrett-et-al-2007}, which only covers
datatypes.
To our knowledge, our procedure is the first of its kind for the theory of
codatatypes.
\begin{conf}
Detailed proofs of soundness and completeness are included in a technical report
\cite{our-report}.
\end{conf}%

@@

Second, \emph{a uniqueness rule replaces the acyclicity rule of datatypes.}
Cyclic constraints such as
$\const{x} \teq \const{S}(\const{x})$ %, where $\const{C}$ is a constructor,
are unsatisfiable for datatypes but satisfiable for codatatypes.
However, the uniqueness principle states that two bisimilar values---i.e., two values
having the same possibly infinite expansion---must be equal; from $\const{x}
\teq \const{S}(\const{x})$ and
$\const{y} \teq \const{S}(\const{y})$, it deduces $\const{x} \teq \const{y}$.
The acyclicity and uniqueness rules are needed to ensure completeness on
ground problems. %and the absence of spurious models.
They cannot be expressed as finite axiomatizations, so they naturally belong in
a decision procedure.
%    * in particular, acyclicity and uniqueness are necessary for some proofs
%%      (and cannot be axiomatized finitely)
%    * and for model finding, without them we quickly get spurious models
%    * explain how finite model finding works
%(For the other (co)datatype properties---the injectivity, distinctness, and
%exhaustiveness of constructors and the selector laws---it is
%widely recognized that decision procedures can be more efficient than
%axiomatizations.)
%axiom.

@@

The last precondition is admittedly rather technical. Its role is to ensure that
$\ValC$ remains injective modulo $\alpha$-equivalence. A simpler, less efficient
way to proceed would be to compute $\nf{\ValC [ {\Varec{x}} \mapsto \MU {\Varec{x}}.\; t ]}$ optimistically and check
injectivity afterward. Injectivity is important to satisfy disequality constraints.
For infinite codatatypes, it is always possible to find suitable values of $v$.

\begin{examplex}
Let $\delta$ be a codatatype with the constructors $\const{C},\, \const{D},\, \const{E} :\delta \rightarrow \delta$.
Let $\Ec$ be the set
$\{
u \teq \const{C}( z ),\;
v \teq \const{D}( z ),\;
w \teq \const{E}( y ),\;
x \teq \const{C}( v ),\;
z \not\teq v
\}$.
After applying the calculus to saturation on $\Ec$,
the mapping $\Val$ is as follows:
%\[\begin{array}{r@{}c@{}l@{\;}l@{\qquad}r@{}c@{}r@{\;}l@{}l@{\;}l}
%\[\begin{array}{r@{}c@{}l@{\;}l@{\qquad}r@{}c@{}r@{\;}l@{}l@{\;}l@{\qquad}r@{}c@{}r@{\;}l}
\[\begin{array}{r@{}c@{}r@{}l@{\qquad}r@{}c@{}r@{}l@{\qquad}r@{}c@{}l}
\Val \ec{u} & {}\,=\,{} & \MU {\Varec{u}}.\; & \const{C}( {\Varec{z}} ) &
\Val \ec{w} & = & \MU {\Varec{w}}.\; & \const{E}( {\Varec{y}} ) &
\Val \ec{y} & {}\,=\,{} & {\Varec{y}} \\
\Val \ec{v} & = & \MU {\Varec{v}}.\; & \const{D}( {\Varec{z}} ) &
\Val \ec{x} & {}\,=\,{} & \MU {\Varec{x}}.\; & \const{C}( \MU {\Varec{v}}.\; \const{D}( {\Varec{z}} ) ) &
\Val \ec{z} & = & {\Varec{z}}
\end{array}\]
%
To construct a completion $\ValC$, we must choose values for the free
variables of $\Val$, specifically ${\Varec{y}}$ and ${\Varec{z}}$.
Modulo $\alpha$-equivalence, $\smash{\Terms^{\smash{\J}}_{{\Varec{z}}}( \Val )} =
%\begin{array}{c}
\{
\MU a.\: \const{C}( a ),\:
\MU a.\: \const{D}( a ),\allowbreak\:
\MU a.\: \const{C}( \const{D}( a )),\allowbreak\:
\const{C}( \MU a.\: \const{D}( a ))
\}
%\end{array}
$.
Consider a normal interpretation $\J$ that evaluates variables in $\Ec$ based on the mapping $\Val$:
$z$ is interpreted as $\Val \ec{z}$, $y$ as $\Val \ec{y}$, and so on.
Assigning a value for $\Val \ec{z}$ that is $\alpha$-equivalent to a value in $\Terms^{\smash{\J}}_{{\Varec{z}}}( \Val )$
may cause values in the range of $\Val$ to become $\alpha$-equivalent,
which in turn may cause $\Ec$ to be falsified by $\J$.
For example, assign $\MU {\Varec{z}}.\; \const{D}( {\Varec{z}} )$ for ${\Varec{z}}$.
After the substitution, $\Val \ec{v} = \MU {\Varec{v}}.\; \const{D}( \MU {\Varec{z}}.\; \const{D}( {\Varec{z}} ) )$,
which has normal form $\MU {\Varec{v}}.\; \const{D}( {\Varec{v}} )$,
which is $\alpha$-equivalent to $\MU {\Varec{z}}.\; \const{D}( {\Varec{z}} )$.
However, this contradicts the disequality $z \not\teq v$ in~$\Ec$.
On the other hand, if the value assigned to ${\Varec{z}}$ is fresh, %not $\alpha$-equivalent to any term in $\Terms^{\smash{\J}}_{{\Varec{z}}}( \Val )$,
the values in the range of $\Val$ remain $\alpha$-disequivalent.
For example, we may assign a value such as
$\MU {\Varec{z}}.\; \const{E}( {\Varec{z}} )$ or
$\MU {\Varec{z}}.\; \const{D}( \const{C}( {\Varec{z}} ) )$
to ${\Varec{z}}$.
\begin{rep}Legal substitutions for ${\Varec{z}}$ may cause the range of $\Val$ to contain abnormal terms.
For example, after assigning $\MU {\Varec{z}}.\; \const{D}( \const{C}( {\Varec{z}} ) )$ to $\Varec{z}$, we have
$\Val \ec{u} = \MU {\Varec{u}}.\; \const{C}( \MU {\Varec{z}}.\; \const{D}( \const{C}( {\Varec{z}} ) ) )$,
with normal form
$\MU {\Varec{u}}.\; \const{C}( \MU {\Varec{z}}.\; \const{D}( {\Varec{u}} ) )$.\end{rep}
\xend
\end{examplex}

>>>>

Codatatypes are freely generated by their constructors, but in contrast with datatypes,
infinit\-e constructor terms are also legitimate values for codatatypes
(Section~\ref{sec:the-theory-of-co-datatypes}). Intuitively, the
values of a codatatype consist of all well-typed finite and infinite ground
constructor
terms, and only those. For example, the coinductive specification
%
\[\keyw{codatatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat})\]
%
(using an ML-like syntax) introduces a type that
models the natural numbers $\const{Z}$, $\const{S}(\const{Z})$, $\const{S}(\const{S}(\const{Z}))$, $\ldots$\afterLdots{},
using Peano notation but extended with an
infinite value $\infty = \const{S}(\const{S}(\const{S}(\ldots)))$.
The equation $\const{S}(\infty) \teq \infty$ holds as expected,
because both sides expand to the infinite term
$\const{S}(\const{S}(\const{S}(\ldots)))$, which uniquely identifies the
value~$\infty$.
Compared with the conventional inductive definition
\begin{rep}
\[\keyw{datatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat}) \,\mid\, \const{Infty}\]
\end{rep}
with a dedicated constructor for representing infinity, the codatatype avoids
one case by unifying the finite and infinite nonzero cases.

Datatypes and codatatypes are an integral part of many proof assistants,
including Agda, Coq, Isabelle, Matita, and PVS. In recent years, datatypes
have emerged in a few automatic theorem provers as well. The SMT-LIB
%\cite{barrett-et-al-2010}
format, implemented by most SMT
%(satisfiability modulo theories)
solvers, has been extended with a syntax for datatypes.
In this \thewordpaper, we introduce a unified decision procedure for ground
problems involving datatypes and codatatypes in
\begin{conf}\goodbreak\noindent\end{conf}%  %%% TYPESETTING: terrible HACK
combination (Section~\ref{sec:the-ground-decision-procedure}).
The procedure is described abstractly as a calculus and can be
composed via the Nelson--Oppen method \cite{nelson-oppen-1979}.
It generalizes the procedure by Barrett et al.\ \cite{barrett-et-al-2007}, which only covers
datatypes.
To our knowledge, our procedure is the first of its kind for the theory of
codatatypes.
\begin{conf}
Detailed proofs of soundness and completeness are included in a technical report
\cite{our-report}.
\end{conf}%

@@

Second, \emph{a uniqueness rule takes the place of the acyclicity rule of datatypes.}
Cyclic constraints such as
$\const{x} \teq \const{S}(\const{x})$ %, where $\const{C}$ is a constructor,
are unsatisfiable for datatypes but satisfiable for codatatypes.
However, the uniqueness principle states that two bisimilar values---i.e., two values
having the same possibly infinite expansion---must be equal; from $\const{x}
\teq \const{S}(\const{x})$ and
$\const{y} \teq \const{S}(\const{y})$, it deduces $\const{x} \teq \const{y}$.
The acyclicity and uniqueness rules are needed to ensure completeness on
ground problems. %and the absence of spurious models.
They cannot be expressed as finite axiomatizations, so they naturally belong in
a decision procedure.
%    * in particular, acyclicity and uniqueness are necessary for some proofs
%%      (and cannot be axiomatized finitely)
%    * and for model finding, without them we quickly get spurious models
%    * explain how finite model finding works
%(For the other (co)datatype properties---the injectivity, distinctness, and
%exhaustiveness of constructors and the selector laws---it is
%widely recognized that decision procedures can be more efficient than
%axiomatizations.)
%axiom.

@@

The last precondition is admittedly rather technical. Its role is to ensure that
$\ValC$ remains injective modulo $\alpha$-equivalence. A simpler, less efficient
way to proceed would be to compute $\nf{\ValC [ {\Varec{x}} \mapsto \MU {\Varec{x}}.\; t ]}$ optimistically and check
injectivity afterward. Injectivity is important to satisfy disequality constraints.
For infinite codatatypes, it is always possible to find suitable values $v$. For example,
if the longest cycle in $\Terms^{\smash{\J}}_{\Varec{x}}(\ValC)$ is of length $n$, a fresh value
can be constructed with a cycle of length $n + 1$.

\begin{examplex}
Let $\delta$ be a codatatype with the constructors $\const{C},\, \const{D},\, \const{E} :\delta \rightarrow \delta$.
Let $\Ec$ be the set
$\{
u \teq \const{C}( z ),\;
v \teq \const{D}( z ),\;
w \teq \const{E}( y ),\;
x \teq \const{C}( v ),\;
z \not\teq v
\}$.
After applying the calculus to saturation on $\Ec$,
the mapping $\Val$ is as follows:
%\[\begin{array}{r@{}c@{}l@{\;}l@{\qquad}r@{}c@{}r@{\;}l@{}l@{\;}l}
%\[\begin{array}{r@{}c@{}l@{\;}l@{\qquad}r@{}c@{}r@{\;}l@{}l@{\;}l@{\qquad}r@{}c@{}r@{\;}l}
\[\begin{array}{r@{}c@{}r@{}l@{\qquad}r@{}c@{}r@{}l@{\qquad}r@{}c@{}l}
\Val \ec{u} & {}\,=\,{} & \MU {\Varec{u}}.\; & \const{C}( {\Varec{z}} ) &
\Val \ec{w} & = & \MU {\Varec{w}}.\; & \const{E}( {\Varec{y}} ) &
\Val \ec{y} & {}\,=\,{} & {\Varec{y}} \\
\Val \ec{v} & = & \MU {\Varec{v}}.\; & \const{D}( {\Varec{z}} ) &
\Val \ec{x} & {}\,=\,{} & \MU {\Varec{x}}.\; & \const{C}( \MU {\Varec{v}}.\; \const{D}( {\Varec{z}} ) ) &
\Val \ec{z} & = & {\Varec{z}}
\end{array}\]
%
To construct a completion $\ValC$, we must choose values for ${\Varec{y}}$ and ${\Varec{z}}$,
which are free in~$\Val$.
Modulo $\alpha$-equivalence, $\smash{\Terms^{\smash{\J}}_{{\Varec{z}}}( \Val )} =
%\begin{array}{c}
\{
\MU a.\: \const{C}( a ),\:
\MU a.\: \const{D}( a ),\allowbreak\:
\MU a.\: \const{C}( \const{D}( a )),\allowbreak\:
\const{C}( \MU a.\: \const{D}( a ))
\}
%\end{array}
$.
Consider a normal interpretation $\J$ that evaluates variables in $\Ec$ based on~$\Val$:
$\interp{\J}{u} = \Val \ec{u}$, $\interp{\J}{v} = \Val \ec{v}$, and so on.
Assigning a value for $\Val \ec{z}$ that is $\alpha$-equivalent to a value in $\Terms^{\smash{\J}}_{{\Varec{z}}}( \Val )$
may cause values in the range of $\Val$ to become $\alpha$-equivalent,
which in turn may cause $\Ec$ to be falsified by $\J$.
For example, assign $\MU {\Varec{z}}.\; \const{D}( {\Varec{z}} )$ for ${\Varec{z}}$.
After the substitution, $\Val \ec{v} = \MU {\Varec{v}}.\; \const{D}( \MU {\Varec{z}}.\; \const{D}( {\Varec{z}} ) )$,
which has normal form $\MU {\Varec{v}}.\; \const{D}( {\Varec{v}} )$,
which is $\alpha$-equivalent to $\MU {\Varec{z}}.\; \const{D}( {\Varec{z}} )$.
However, this contradicts the disequality $z \not\teq v$ in~$\Ec$.
On the other hand, if the value assigned to ${\Varec{z}}$ is fresh, %not $\alpha$-equivalent to any term in $\Terms^{\smash{\J}}_{{\Varec{z}}}( \Val )$,
the values in the range of $\Val$ remain $\alpha$-disequivalent.
We can assign a value such as
$\MU {\Varec{z}}.\; \const{E}( {\Varec{z}} )$,
$\MU {\Varec{z}}.\; \const{D}( \const{C}( {\Varec{z}} ) )$, or
$\MU {\Varec{z}}.\; \const{C}( \const{C}( \const{D}( {\Varec{z}} ) ) )$
to ${\Varec{z}}$.
\begin{rep}Legal substitutions for ${\Varec{z}}$ may cause the range of $\Val$ to contain abnormal terms.
For example, after assigning $\MU {\Varec{z}}.\; \const{D}( \const{C}( {\Varec{z}} ) )$ to $\Varec{z}$, we have
$\Val \ec{u} = \MU {\Varec{u}}.\; \const{C}( \MU {\Varec{z}}.\; \const{D}( \const{C}( {\Varec{z}} ) ) )$,
with normal form
$\MU {\Varec{u}}.\; \const{C}( \MU {\Varec{z}}.\; \const{D}( {\Varec{u}} ) )$.\end{rep}
\xend
\end{examplex}
